\section{Introduction}
\label{sec:intro}

In his iconic quote, referenced in the title of this thesis, the famous 16\textsuperscript{th}-century Polish poet Mikołaj Rej declared his conviction that Polish people have their own language: 

\begin{quote}
\centering
   \textit{A niechaj narodowie wżdy postronni znają, \\
    Iż P o l a c y nie gęsi, iż swój język mają!}\footnote{ `And may the other nations finally know that Poles are not geese, that they have their own language!'} \\
\raggedleft
\citet{mikołaj_rej}\footnote{While the reference is to a 2015 edition of the poet's collected works, the poem itself was written in 1562.}
\end{quote}

While one would be hard-pressed to find someone trying to dispute the existence of the Polish language, one perhaps should consider certain alterations to the quote itself: the average Polish user no longer speaks in the same fashion as Mikołaj Rej did when he wrote his poetry, and variation occurs not only diachronically, but also regionally. Perhaps it would be more correct to say then that Poles \textit{swe języki mają!}\footnote{ `have their own languages!'}

Language variation plays an essential role in natural language processing: natural language as used by speakers is ever-changing, and NLP tools have to, to some extent, account for even the synchronic variation. As \citet{Zampieri2020NaturalLP} or \citet{dorn-2019-dialect} point out, developing methods to handle language variation is also relevant for adapting the existing tools to minority languages or dialects. While there are methods that allow for the utilization of raw data, \citet{ponti_2019} remarks that other methods still rely on annotated data which is difficult to come by for less popular languages. Simultaneously, as \citet{quantitative-historical} note, historical linguistics on its own is a data-driven field, and access to data, as well as methods for processing it, are very important; at the same time, they highlight the usefulness of annotated corpora. This annotation may be especially useful when it comes to languages with a richer morphology, such as Slavic languages, as it may enable e.g. searching for all the inflectional forms of a given word \citep{pęzik_2012}. Thus, identifying the kinds of variation that occur between two languages or dialects not only constitutes a contribution to the body of knowledge about those languages on its own but also opens up possibilities for adapting existing tools for major languages to be used for the partial automatizing of data annotation. 

\subsection{Research Questions}
\label{subsec:research-questions}

The inquiries conducted within this thesis are based on a 19\textsuperscript{th}-century memoir by Juliusz Czermiński, who came from the area of nowadays Eastern Poland and Western Ukraine. Given its age as well as the sociopolitical and geographical context, the text is expected to differ from standard modern Polish, and to bear features typical for the historical dialect of that area. The data itself and the annotation process (conducted by the author of the thesis) are outlined in more detail in \autoref{subsec:data} and \autoref{subsec:annotation}. With the previously mentioned issues of language variation both in NLP and in historical linguistics, and the research described in \autoref{sec:background}, the hope is for this text to yield some insights concerning possibilities for the identification of language variation in Polish using computational tools and resources. 

Consequently, the aims of this thesis can be described as two sides of the same coin, as it seeks to simultaneously answer the following two questions:
\begin{enumerate}
    \item \textit{Is it possible to identify language variation in terms of orthography, morphology, and syntax in a Polish text using tools and resources such as lemmatizers, POS-taggers, and modern corpora?}
    \item \textit{In what ways does the text in question, a 19\textsuperscript{th}-century memoir by Juliusz Czermiński, differ from modern standard Polish?}
\end{enumerate}

\subsection{Motivation}
\label{subsec:motivation}

As mentioned at the start of this section, investigating language variation is not only its own field of linguistics but is also relevant for NLP --- and methods and discoveries within these fields can inform each other. It is interesting to see how tools intended for working with modern languages can be used to identify ways in which those differ from their historical counterparts. These differences can help inform the pre-processing of the texts or ways in which the tools need to be adapted to enable a more reliable data annotation, which, in turn, can be used for further qualitative, corpus-based inquiries into some historical form of a language. 

Simultaneously, language change is not only a thing of the past, but a continuous process that may lead to modern tools becoming outdated in the future; furthermore, language varies also based on factors other than time, such as geography or social class, and this variation can also prove problematic to NLP applications, e.g. when it comes to language recognition or processing of historical data \citep{jurgens-etal-2017-incorporating, regnault-etal-2019-challenges, Zampieri2020NaturalLP}. While methods such as cross-linguistic transfer learning do exist, they cannot be applied to all the tools equally, especially non-neural ones or ones that are not available for fine-tuning.

While the major focus of this thesis is on exploring historical and dialectical language variation in Polish alongside the methods that can be employed for such investigations, hopefully, it can yield insights into and spark some discussion about related topics, such as the handling of linguistic variation in NLP, computational methods in historical linguistics, as well as resources for historical linguistics and the annotation thereof.

\subsection{Contributions}
\label{subsec:contributions}

Within this thesis, a variety of ways for discovering historical linguistic variation using tools intended for modern languages is tested. The methods are reviewed with regard to the kind of results that they yield and the amount of annotation or preparation needed. While the majority of the methods do require the data to be annotated, a number of observations (described in more detail in \autoref{sec:results}) can be made based on the results, such as the influence of nonstandard capitalization on tagging and lemmatizing, prevalent spelling differences reflecting a different spelling standard or different pronunciation, issues that the existing tools have with various kinds of proper names; additionally, certain gender bias is revealed to exist as far as the XPOS tagset and the tools used to annotate with it, which can hopefully contribute to the discussion concerning the tagsets used for annotating data in Polish and certain biases that are possibly present in the training data as well.

Simultaneously, due to the nature of the experiments, a second set of observations emerges, as a comparison of the tagging and lemmatizing performance of multiple tools on modern data is conducted, which can inform the choice of a tool for other research or real-life application. Finally, the data that has been manually annotated by the author of this project and that is utilized within it constitutes a non-negligible contribution to the body of annotated historical data for Polish. 

\subsection{Scope}
\label{subsec:scope}

This thesis encompasses attempting to identify language variation in a specific text with the use of tools such POS taggers and lemmatizers, resources such as corpora, and methods such as n-gram count analysis. These are used to approximate variation in terms of spelling, vocabulary, and syntax, although certain observations concerning the pronunciation (the consistent shift from /a/ to /ɛ/) are also made. However, due to the data in question coming from a single author, the conclusions cannot be extended to the language of the time and region at large.